---
title: "wheather: Scoring Methodology"
subtitle: "How we measure whether the weather was actually better"
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
---

```{r}
#| label: setup
#| include: false
library(wheather)
library(data.table)
library(ggplot2)
```

## The Problem

"This summer's weather was way better than last year's." Everyone has an opinion, but how do you actually measure it? **wheather** turns daily weather data into a single 0-100 "goodness" score and compares periods statistically.

## Data Source

We use the [OpenWeather One Call API 3.0](https://openweathermap.org/api/one-call-3) Day Summary endpoint, which provides daily aggregates back to 1979:

- Temperature: min, max, morning, afternoon, evening, night (°C)
- Precipitation: total (mm)
- Cloud cover: afternoon (%)
- Humidity: afternoon (%)
- Wind: max speed (m/s)

## Scoring Components

Each day gets five component scores (0-100), then a weighted average produces the total score.

### Temperature Score (default weight: 30%)

A bell curve centred on an "ideal range" (default 22-26°C, configurable):

$$
\text{score} = \max\left(0,\; 100 - r \cdot d^{p}\right)
$$

Where $d$ = distance from the nearest ideal bound, $r$ = decay rate (default 5), $p$ = decay power (default 1.5).

```{r}
#| label: temp-curve
#| fig-cap: "Temperature scoring curve (default parameters)"
params <- default_params()
temps <- seq(5, 45, by = 0.5)
scores <- vapply(temps, function(t) {
  score_temp(t, NA, NA, params)
}, numeric(1))

ggplot(data.frame(temp = temps, score = scores), aes(temp, score)) +
  geom_line(linewidth = 1.2, colour = "#2980b9") +
  geom_vline(xintercept = c(22, 26), linetype = "dashed", alpha = 0.5) +
  annotate("rect", xmin = 22, xmax = 26, ymin = 0, ymax = 100,
           fill = "#2ecc71", alpha = 0.15) +
  labs(x = "Afternoon Temperature (°C)", y = "Score") +
  theme_minimal(base_size = 14)
```

### Rain Score (default weight: 20%)

Exponential decay — dry days score highest:

$$
\text{score} = 100 \cdot e^{-k \cdot \text{rain\_mm}}
$$

Where $k$ is calibrated so that 10mm (heavy rain) scores ~15.

```{r}
#| label: rain-curve
#| fig-cap: "Rain scoring curve"
rain_mm <- seq(0, 30, by = 0.5)
scores <- vapply(rain_mm, function(r) score_rain(r, params), numeric(1))

ggplot(data.frame(rain = rain_mm, score = scores), aes(rain, score)) +
  geom_line(linewidth = 1.2, colour = "#3498db") +
  labs(x = "Total Precipitation (mm)", y = "Score") +
  theme_minimal(base_size = 14)
```

### Cloud Cover Score (default weight: 20%)

Linear: $\text{score} = 100 - \text{cloud\_cover\_pct}$

### Humidity Score (default weight: 15%)

Comfort zone at 30-60%, linear decay outside.

### Wind Score (default weight: 15%)

Pleasant up to 15 km/h, linear decay to 0 at 50 km/h.

## Weighting

The default weights reflect a general "good weather = warm, sunny, dry" preference:

| Component | Weight |
|-----------|--------|
| Temperature | 30% |
| Rain | 20% |
| Cloud cover | 20% |
| Humidity | 15% |
| Wind | 15% |

These are fully configurable. Someone who hates wind might set wind to 30% and reduce temperature.

## Statistical Comparison

Given two scored periods, we compute:

1. **Mean and SD** of daily scores for each period
2. **Welch's t-test** — is the difference statistically significant?
3. **Cohen's d** — how large is the effect?
   - d < 0.2: negligible
   - 0.2-0.5: small
   - 0.5-0.8: medium
   - d > 0.8: large

## Example: Sydney Summer 2024-25 vs 2023-24

*(Run this with your API key to see real results)*

```{r}
#| label: example
#| eval: false
result <- compare_periods(
  lat = -33.87, lon = 151.21,
  start1 = "2024-12-01", end1 = "2025-02-28",
  start2 = "2023-12-01", end2 = "2024-02-29"
)
print(result)
```

## Limitations

- **"Good weather" is subjective** — the scoring reflects configurable preferences, not objective truth. A surfer and a hiker score the same day differently.
- **Daily aggregation loses nuance** — a day with morning rain and afternoon sun gets averaged.
- **No UV or air quality** — the Day Summary endpoint doesn't include these.
- **Comparison assumes similar day counts** — comparing 90 days to 30 days works statistically but may feel unfair.
